---
title: "Viperfish Overview"
author: "Aaron Robotham"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Viperfish Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## What Viperfish Does

**Viperfish** is a post-processing package that turns semi-analytic galaxy model outputs
into synthetic observables. In practical terms, it takes star-formation and metallicity
histories from **Shark** and/or galaxy mappings from **Stingray**, and computes
spectral energy distributions (SEDs), broad-band magnitudes, and optional spectra
using **ProSpect**.

This fills the final step between physically-motivated galaxy evolution predictions and
survey-facing mock products.

## Why This Package Exists

A common simulation workflow is:

1. Dark-matter simulations and halo/subhalo catalogues are prepared.
2. **Shark** evolves galaxy baryonic components in those merger trees and writes per-snapshot outputs.
3. **Stingray** projects those outputs into light-cone/mock-survey geometry.
4. **Viperfish** maps each galaxy history to synthetic photometry/spectra suitable for science analyses.

Without this final SED synthesis step, mocks are still effectively "dark" with respect to
many observational selection functions and derived photometric products.

## Connection to Shark Data Products

Viperfish directly interfaces with Shark's HDF5 schema for star-formation histories,
notably:

- `bulges_diskins/star_formation_rate_histories`
- `bulges_mergers/star_formation_rate_histories`
- `disks/star_formation_rate_histories`
- corresponding `metallicity_histories`
- `galaxies/id_galaxy`
- `lbt_mean`
- cosmology metadata including `cosmology/h`

In Shark documentation, these SFR histories are stored in **Msun/yr/h**. Viperfish converts
them to physical units before evaluating spectra, which is essential for physically consistent
luminosity and magnitude outputs.

Viperfish also supports optional per-galaxy extinction files that can be read and applied
at generation time.

## Core Interfaces

### `genShark`: snapshot/subvolume processing

Use `genShark` when you want to process one Shark snapshot/subvolume directly.

- reads SFH and metallicity histories
- computes bulge/disk/total outputs
- returns in-memory products and can optionally write files

### `genSting`: light-cone processing

Use `genSting` when starting from a Stingray light-cone. It:

- reads the light-cone galaxy mapping
- groups rows by `(snapshot, subvolume)`
- loads matching Shark histories on-demand
- computes photometry or spectra in parallel
- optionally writes final outputs after temporary chunked processing

This design keeps memory use practical for large mock catalogues.

### `genSED`: lower-level engine

`genSED` is the lower-level interface consumed by both high-level pathways. It combines
three star-formation channels (bulge from disk instabilities, bulge from mergers, disk),
applies attenuation/re-emission assumptions, and computes output magnitudes/spectra.

## Physical Modelling Choices

Viperfish follows a pragmatic modelling setup intended for scalable mock generation:

- stellar population libraries via ProSpect (e.g. BC03/EMILES/BPASS where available)
- Charlot & Fall style attenuation controls (`tau_*`, `pow_*`)
- Dale-style dust re-emission controls (`alpha_SF_*`)
- optional radio continuum contributions for star-formation

The goal is not to encode every physically possible dust geometry, but to provide a
repeatable and computationally tractable baseline that can be used at survey scale.

## Output Products

In photometric mode, Viperfish returns component-level magnitudes and IR diagnostics,
including bulge/disk/total variants for intrinsic and dust-processed quantities.

In spectrum mode, Viperfish returns wavelength/flux outputs over a selected range and
supports downstream re-binning utilities for final catalog products.

Output writing supports CSV and HDF5 (format inferred from extension where applicable).

## Typical Usage Pattern

```{r eval = FALSE}
# Shark snapshot/subvolume workflow
sed_box <- genShark(
  path_shark = "/path/to/shark/output",
  snapshot = 199,
  subvolume = 0,
  write_final_file = TRUE,
  final_file_output = "Shark-SED.csv"
)

# Stingray light-cone workflow
sed_lc <- genSting(
  file_sting = "/path/to/mocksurvey.hdf5",
  path_shark = "/path/to/shark/output",
  write_final_file = TRUE,
  final_file_output = "Stingray-SED.csv"
)
```

## Practical Guidance

- Keep `h='get'` unless you have a strong reason not to.
- Use default filters as a robust baseline, then extend as needed.
- Start with photometric mode for catalogue production; use spectrum mode for targeted studies.
- For very large runs, use `genSting(..., restart=TRUE)` to recover from interrupted jobs.
- Prefer writing temporary/final outputs to fast local storage.

## Summary

Viperfish is the bridge from Shark/Stingray structural galaxy predictions to observables.
It provides a high-level, parallel-aware interface that turns SAM outputs into physically
interpretable SED products for mock survey science, calibration experiments, and
methodology development.
